---
title: "Class1 for R"
author: "Patrick Shema - 101375"
date: "2025-10-16"
output: html_document
---

Vector: The foundational R data structure
 Numeric
```{r}
x <- c(0.5, 0.6)
age<-c(20,35,32,29)
summary(age)
 
age<-c(35,24,18,24)
 mean(age)
```
 
 Logical 
```{r}
 x <- c(TRUE, FALSE)
 x <- c(T, F)

```
 
 
Character
```{r}
x<-c("a","b","c")
 class<-c("M","F","F","M")
```
 
Integer
```{r}
x<-9:29
x
```

Complex
```{r}
 x<-c(1+0i,2+4i)
```

Matrixandoperationwithinmatrices
```{r}
 x<-matrix(1:6,nrow=2,ncol=3)
 x
```

```{r}
(y<-matrix(1:6,nrow=3,ncol=2))
```
```{r}
x %*% y
```

List
```{r}
  x<-list(1,"a",TRUE,1+4i)
 x
```

Factor
```{r}
  x<-factor(c("yes","yes","no","yes","no"))
 x
```

Create a vector with NAs init
```{r}
 x<-c(1,2,NA,10,3)

```


ReturnalogicalvectorindicatingwhichelementsareNA
```{r}
 is.na(x)
```
```{r}
x<-c(1,2,4,"NA",5)
 bad<-is.na(x)
 print(bad)
 bad
```
```{r}
 x[!bad]
```

What if there are multiple R objects and you want to take the subset with no missing valuesinanyofthoseobjects?
```{r}
x<-c(1,2,NA,4,NA,5)
 y<-c("a","b",NA,"d",NA,"f")
 good<-complete.cases(x,y)
 good
```

Coercion
If character is present,in a vector, R convert everything in the vector to character strings. If a vector only contains logical and numbers,R will convert the logical to numbers,Every true becomes a 1,and every FALSE becomes 0
```{r}
 sum(c(TRUE,TRUE,FALSE,FALSE,FALSE))
```

Createdata

Create data Using datframe. Dataframe is more general than matrix How???
Because dataframe can contain different modes of data(Numeric,character and so on)Similar to what you can see in SPSS,SAS,…
 ###Let’screateadataframe
 
```{r}
 studentID<-c(1,2,3,4,5)
 math_score<-c(12,17,10,9,NA)
 gender<-c("M","F","M","M","F")
 it_score<-c(13,18,11,10,19)
 scoredata<-data.frame(studentID,gender,math_score,it_score)
 scoredata

```
 
Create data from keyboard(ormatrix) with variable names2. Invoke the text editor in the data objected created at first step.
```{r}
#data_class2<-data.frame(height=numeric(0),weight=numeric(0),bmi=numeric(0)
#data_class2<-edit(data_class2)
HEIGHT<- c(161,155,158,170,160,162,158,158,167,160,155,154,155,157,157,160,158,160,160,152,154,150,161,162,164,164,161,155,159,163,159,160,158,165,156,163,155,164,163,168,157,164)

WEIGHT<- c(50,49,42,65,60,52,58,46,45,51,60,42,53,48,48,48,53,52,51,53,44,56,63,52,57,49,52,54,46,50,61,55,45,63,60,56,52,47,52,55,48,58)

data_class<- data.frame(HEIGHT,WEIGHT)
data_class
```


SummaryStatistics 
```{r}

# summary(data_class)
# apply(data_class,2,mean)
# apply(data_class,2,sd)
# c(mean(data_class$HEIGHT),sd(data_class$HEIGHT))
# c(mean(data_class$WEIGHT),sd(data_class$WEIGHT))
# c(Mean=mean(data_class$HEIGHT),SD=sd(data_class$HEIGHT))
# c(Mean=mean(data_class$WEIGHT),SD=sd(data_class$WEIGHT))
```
 
 
Variation and covariation using both numerical and graphs

Relationship between a categorical and a continuous variable

```{r}
data("iris")
head(iris)
```
```{r}
mean(iris$Petal.Length)
```


```{r}
mean(iris$Petal.Length[iris$Species=="setosa"])
```

```{r}
mean(iris$Petal.Length[iris$Species=="versicolor"])
```
```{r}
mean(iris$Petal.Length[iris$Species=="virginica"])
```

```{r}
##shortcut
by(iris$Petal.Length,iris$Species,mean)
```

```{r}
by(iris$Petal.Length,iris$Species,sd)
```
```{r}
by(iris$Petal.Length,iris$Species,summary)
```

Visualize the differences in continuous variables between categories using Box-and whisker plot

```{r}
 boxplot(iris$Sepal.Length~iris$Species,data=iris, main="Comparison")
```

 Relationship between two categorical variables

```{r}
ucba<-data.frame(UCBAdmissions)
 head(ucba)
```

```{r}
cross<-xtabs(Freq~Gender+Admit,data=ucba)
(cross<-xtabs(Freq~Gender+Admit,data=ucba))
```

```{r}
 ##IstheregenderbiasinUCBgraduateadmissionprocess?
 prop.table(cross,2)
```
 Simpson’s paradox
 
 Phenomenon,where a trend that appears in combined groups of data disappears or reverses when broken down into groups.
 
```{r}
cross2<-xtabs(Freq~Gender+Admit,data=ucba[ucba$Dept=="A",])
 prop.table(cross2,1)
```
BarChart

```{r}
 dat<-read.table(text="ProdAProdBProdCProdD
 1110506070
 2120508065",header=TRUE)
 barplot(as.matrix(dat),beside=FALSE,col=c("Red","green"))

```


```{r}
#barplot(as.matrix(dat),beside=TRUE,col=c("gold3","red"))
 dat <- read.table(text = "A B C D E F G
 1 10 80 30 90 70 60 90
 2 20 50 70 50 40 10 40
 3 60 80 80 60 60 30 160
 4 20 40 70 80 20 10 70", header = TRUE)
 barplot(as.matrix(dat))

```

Scatter Plot

```{r}
plot(WEIGHT~HEIGHT,data=data_class)
```

```{r}
plot(data_class$HEIGHT,data_class$WEIGHT)
```

```{r}
plot(data_class[,1],data_class[,2])

```

 Linear Models
```{r}
 plot(WEIGHT~HEIGHT,data=data_class)
 abline(lm(WEIGHT~HEIGHT,data=data_class)$coefficient)
```


```{r}
data_lm<-lm(WEIGHT~HEIGHT,data=data_class)
 summary(data_lm)
```
```{r}
 hist(data_class$HEIGHT)
```


```{r}
hist(data_class$WEIGHT,col = "blue")

```


Linear with other line

```{r}
plot(WEIGHT~HEIGHT,data=data_class)
 abline(lm(WEIGHT~HEIGHT,data=data_class)$coefficient)
 lines(lowess(data_class$HEIGHT,data_class$WEIGHT),col="blue")

```

```{r}
cor(data_class$HEIGHT,data_class$WEIGHT)
```

Data management

 Create another Variable
 
```{r}
attach(data_class)
 (BMI<-WEIGHT/(HEIGHT/100)^2)
```
 

```{r}
(BMI<-round(WEIGHT/(HEIGHT/100)^2,digit=2))
```

```{r}
 head(cbind(data_class,BMI),n=5)
```

```{r}
new_data_class<-cbind(data_class,BMI)
 tail(cbind(data_class,BMI),n=10)
```

 SummaryofBMI
 
```{r}
 summary(BMI)
```
 
```{r}
 stem(BMI,scale=2)
```
 
 